A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-27 paper.tex
Reorder event section
And stick some intermediate headings in to make the logic clearer to ourselves (these can be removed later if need be)

@@

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG.
To recover
the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the lineage carrying the ancestry
to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough
that the number of extant lineages remains
much smaller than $N$ at all times.
While the coalescent model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, which is increasingly
captured in the incredibly rich datasets that are available today.

@@

A final limitation of the eARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts? Gregor: I recall this paper from
Arabidopsis https://www.pnas.org/doi/10.1073/pnas.1211827110]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

Fortunately, it is straightforward to generalise the classical Griffiths
eARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(d,e)$
and the interval $(x, L)$ with $(d, f)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between positions $x$ and $y$ was inherited by
node $d$ from $e$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
$\{(x, y]\}$ with $(d,f)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

>>>>

\subsection*{Node versus edge annotation}

@@

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG.
To recover
the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event above a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, but depending on the ordering of
otherwise indistinguishable objects is generally problematic. In practice,
several methods explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

Another limitation of the eARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts? Gregor: I recall this paper from
Arabidopsis https://www.pnas.org/doi/10.1073/pnas.1211827110]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

Fortunately, it is straightforward to generalise the classical Griffiths
eARG encoding to both simplify implementations and to encompass
alternative patterns of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(d,e)$
and the interval $(x, L)$ with $(d, f)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between positions $x$ and $y$ was inherited by
node $d$ from $e$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
$\{(x, y]\}$ with $(d,f)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

% it's not reasonable to assume that one type of event happens
% at a time any more.
\subsection*{Issues with event-based encoding}

Even with the change to edge annotations, a focus on specific
types of events can limit the types of ancestral history that can
be described by an eARG. As it is derived from the coalescent
with recombination, a standard eARG explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough
that the number of extant lineages remains
much smaller than $N$ at all times.
While the coalescent model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, which is increasingly
captured in the incredibly rich datasets that are available today.

@@

For these reasons, we suggest that instead of defining the nodes in an
ARG as events, which limits us to certain models of inheritance,
we use a more flexible interpretation where they are defined as genomes.
